---
title: "Validation of Grimes' GRIM-U code"
author: "Ian Hussey"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: show
    highlight: haddock
    theme: flatly
    toc: yes
    toc_float: yes
---

```{r include=FALSE}

# formatting options
# set default chunk options
knitr::opts_chunk$set(message = FALSE, 
                      warning = FALSE)

```

# Problem setting

From my understanding, Grimes' simrank() requires an integer input for U. To simulate a half-integer target (e.g., 70.5), the code must simulate the integer above it (e.g., 71) and then force a tie. This tie mathematically subtracts 0.5 from the score, hitting the target ($71 - 0.5 = 70.5$).

However, the original code used `round()`, which follows "Banker's Rounding" (rounding to the nearest even number). eg the following illustrates that base::round() doesn't perform how many people expect:

```{r}
round(0.5)
round(1.5)
round(2.5)
round(3.5)
round(4.5)
round(5.5)
```

If so, this creates a bug in 50% of cases. 

- Failure Case: For a target of 70.5, round(70.5) rounds down to `r round(70.5)`. The tie logic then calculates $70 - 0.5 = 69.5$, which is incorrect.
- Success Case: For a target of 71.5, round(71.5) rounds up to `r round(71.5)`. The tie logic calculates $72 - 0.5 = 71.5$, which works.

The fix is to use `ceiling(k)` instead of `round(k)`. `ceiling()` always rounds up to the next integer, ensuring we always start with the higher value needed for the subtraction logic.

So `j <- simrank(n1, n2, U_target = round(k))` becomes `j <- simrank(n1, n2, U_target = ceiling(k))`

An alternative would be Lukas Jung's excellent {roundwork} package for `round_up()` to produce the expected behavior, but that's another dependency.

# Demonstration of the impact of the change

## Original code 

```{r}

simrank <- function(n1, n2, U_target, max_iter = 100000) {
  total_ranks <- 1:(n1 + n2)
  R1_target <- U_target + n1 * (n1 + 1) / 2
  
  for (i in 1:max_iter) {
    group1_ranks <- sort(sample(total_ranks, n1, replace = FALSE))
    if (sum(group1_ranks) == R1_target) {
      group2_ranks <- setdiff(total_ranks, group1_ranks)
      return(list(
        group1_ranks = group1_ranks,
        group2_ranks = group2_ranks,
        U = U_target
      ))
    }
  }
  
  # Return NA instead of stopping with an error
  return(NA)
}


#Here enter in your n1 and n2
n1 <- 18
n2 <- 60

#Expected stats (Ties and no ties)
uexp <- (n1*n2)/2
seu <- sqrt(n1*n2*(n1+n2+1)/12)
seu_tie <- sqrt(n1*n2*(n1+n2+1)/12 - n1*n2*(7)/(12*(n1 + n2)*(n1 + n2 - 1)) )

#range of p-values of interest
pvals <- seq(0.01, 0.10, by = 0.005)
zscores <- qnorm(1 - pvals/2)

#conversion (no ties)
upred <- seu*zscores + uexp  #has higher max value
upredtie <- seu_tie*zscores + uexp #has lower min value 
vals <- seq(floor(upredtie[length(upredtie)]),ceiling(upred[1]) , by = 0.5)
exactval <- rep(NA_real_, length(vals))
spssval <- rep(NA_real_, length(vals))




#Let it run 
i <- 1
while (i <= length(vals)) {
  k <- vals[i]
  j <- simrank(n1, n2, U_target = round(k))
  
  # Skip iteration if simulation failed
  if (is.na(j[1])) {
    exactval[i] <- NA
    spssval[i] <- NA
    i <- i + 1
    next
  }

  # If k is an integer
  if (round(k) == k) {
    m1 <- wilcox.test(j$group1_ranks, j$group2_ranks)
    m2 <- wilcox.test(j$group1_ranks, j$group2_ranks, correct = FALSE, exact = FALSE)
  } else {
    # Adjust one value to simulate a fractional rank
    w <- which(diff(j$group2_ranks) == 2)[1]
    if (!is.na(w)) {
      mv <- j$group2_ranks[w]
      j$group2_ranks[w] <- mv + 0.5

      w2 <- which(j$group1_ranks == mv + 1)
      if (length(w2) > 0) {
        j$group1_ranks[w2] <- mv + 0.5
      }
    }

    m1 <- wilcox.test(j$group1_ranks, j$group2_ranks)
    m2 <- wilcox.test(j$group1_ranks, j$group2_ranks, correct = FALSE, exact = FALSE)
  }

  exactval[i] <- m1$p.value
  spssval[i] <- m2$p.value

  i <- i + 1
}

n1v <- rep(n1,i-1)
n2v <- rep(n2,i-1)

df <- data.frame("U_Values" = vals, "P_Exact" = exactval, "Approx P" = spssval)
df$Diff <- abs(df$Approx.P - df$P_Exact)
df$n1 <- n1v
df$n2 <- n2v

```

Ian's minor addition to allow comparisons:

```{r}

library(tidyverse)
library(waldo)

df <- df |>
  rename(p_approx = Approx.P,
         p_exact = P_Exact) |>
  as_tibble()

```

## Fixed issue re rounding k

```{r}

simrank <- function(n1, n2, U_target, max_iter = 100000) {
  total_ranks <- 1:(n1 + n2)
  R1_target <- U_target + n1 * (n1 + 1) / 2
  
  for (i in 1:max_iter) {
    group1_ranks <- sort(sample(total_ranks, n1, replace = FALSE))
    if (sum(group1_ranks) == R1_target) {
      group2_ranks <- setdiff(total_ranks, group1_ranks)
      return(list(
        group1_ranks = group1_ranks,
        group2_ranks = group2_ranks,
        U = U_target
      ))
    }
  }
  
  # Return NA instead of stopping with an error
  return(NA)
}


#Here enter in your n1 and n2
n1 <- 18
n2 <- 60

#Expected stats (Ties and no ties)
uexp <- (n1*n2)/2
seu <- sqrt(n1*n2*(n1+n2+1)/12)
seu_tie <- sqrt(n1*n2*(n1+n2+1)/12 - n1*n2*(7)/(12*(n1 + n2)*(n1 + n2 - 1)) )

#range of p-values of interest
pvals <- seq(0.01, 0.10, by = 0.005)
zscores <- qnorm(1 - pvals/2)

#conversion (no ties)
upred <- seu*zscores + uexp  #has higher max value
upredtie <- seu_tie*zscores + uexp #has lower min value 
vals <- seq(floor(upredtie[length(upredtie)]),ceiling(upred[1]) , by = 0.5)
exactval <- rep(NA_real_, length(vals))
spssval <- rep(NA_real_, length(vals))




#Let it run 
i <- 1
while (i <= length(vals)) {
  k <- vals[i]
  j <- simrank(n1, n2, U_target = ceiling(k)) # fixed: round() changed to ceiling()
  
  # Skip iteration if simulation failed
  if (is.na(j[1])) {
    exactval[i] <- NA
    spssval[i] <- NA
    i <- i + 1
    next
  }

  # If k is an integer
  if (round(k) == k) {
    m1 <- wilcox.test(j$group1_ranks, j$group2_ranks)
    m2 <- wilcox.test(j$group1_ranks, j$group2_ranks, correct = FALSE, exact = FALSE)
  } else {
    # Adjust one value to simulate a fractional rank
    w <- which(diff(j$group2_ranks) == 2)[1]
    if (!is.na(w)) {
      mv <- j$group2_ranks[w]
      j$group2_ranks[w] <- mv + 0.5

      w2 <- which(j$group1_ranks == mv + 1)
      if (length(w2) > 0) {
        j$group1_ranks[w2] <- mv + 0.5
      }
    }

    m1 <- wilcox.test(j$group1_ranks, j$group2_ranks)
    m2 <- wilcox.test(j$group1_ranks, j$group2_ranks, correct = FALSE, exact = FALSE)
  }

  exactval[i] <- m1$p.value
  spssval[i] <- m2$p.value

  i <- i + 1
}

n1v <- rep(n1,i-1)
n2v <- rep(n2,i-1)

df_fixed <- data.frame("U_Values" = vals, "P_Exact" = exactval, "Approx P" = spssval)
df_fixed$Diff <- abs(df_fixed$Approx.P - df_fixed$P_Exact)
df_fixed$n1 <- n1v
df_fixed$n2 <- n2v

```

Ian's minor addition to allow comparisons:

```{r}

df_fixed <- df_fixed |>
  rename(p_approx = Approx.P,
         p_exact = P_Exact) |>
  as_tibble()

```

## Comparing original and fixed

```{r}

waldo::compare(df, df_fixed)

```

- this changes p_exact and p_exact values by up to c. .003, and affects a large proportion of cases.

# Other issues found in Grimes' code

In the process of building `{analyticgrimu}` I noticed a handful of further issues in `grim_u.r`. Each is recorded below for the record; the analytic engine in `{analyticgrimu}` (see `R/grimu_map_pvalues.R`) supersedes Grimes' tie/asymptotic handling on every one of these axes.

## Issue: tie-corrected variance uses the constant 7 where the correct value is 6

In `grim_u.r` the tie-corrected SE is computed as:

```r
seu_tie <- sqrt(n1 * n2 * (n1 + n2 + 1) / 12 -
                  n1 * n2 * (7) / (12 * (n1 + n2) * (n1 + n2 - 1)))
```

The standard tie-correction in the Mann–Whitney variance is $\sum_i (t_i^3 - t_i)$ summed over tie groups, where $t_i$ is the size of the $i$-th tie group:

$$\sigma^2 \;=\; \frac{n_1 n_2}{12}\left((N+1) \;-\; \frac{\sum_i (t_i^3 - t_i)}{N(N-1)}\right)$$

For one tied pair (the only tie pattern Grimes' code injects; see the next issue), $t = 2$ gives $\sum (t^3 - t) = 2^3 - 2 = 6$, not 7. Moreover, no tie partition can ever produce 7, because $t^3 - t = t(t-1)(t+1)$ is always divisible by 6 (three consecutive integers always include a multiple of 2 and a multiple of 3). The constant 7 in `seu_tie` is therefore an off-by-one bug/typo: the formula sits half-way between the no-tie correction (0) and the smallest achievable correction (6).

```{r}
# t^3 - t for t = 2, 3, 4, 5, 6, 7 — none equals 7
t <- 2:7
data.frame(t = t, "t_cubed_minus_t" = t^3 - t)
```

`{analyticgrimu}` corrects this by enumerating all achievable values of $\sum (t_i^3 - t_i)$ via `achievable_K_values()`, and 6 (rather than 7) is the second entry in that sequence.

## Issue: at most one tied pair is ever injected

The tie-injection block in `grim_u.r` finds *one* place where two consecutive ranks differ by 2 and rewrites both ranks to a shared half-integer value:

```r
w <- which(diff(j$group2_ranks) == 2)[1]
if (!is.na(w)) {
  mv <- j$group2_ranks[w]
  j$group2_ranks[w] <- mv + 0.5
  w2 <- which(j$group1_ranks == mv + 1)
  if (length(w2) > 0) {
    j$group1_ranks[w2] <- mv + 0.5
  }
}
```

This corresponds to exactly one $t = 2$ tied pair, and only when the simulated rank pattern has a suitable adjacent gap. Real data routinely contains *multi-pair* tie patterns (e.g. two or more separate pairs), *larger* tie groups ($t = 3, 4, \ldots$), and combinations thereof, each of which yields a different $\sum (t_i^3 - t_i)$ and therefore a different $\sigma$ and asymptotic $p$. None of these are modelled in `grim_u.r`.

`{analyticgrimu}::grimu_map_pvalues()` enumerates the full family of achievable K values for partitions with parts ≥ 2 and sum ≤ min(N, 15), producing one (corrected, uncorrected) asymptotic $p$-value column per K. This is the family of values reachable by R `wilcox.test()`, SPSS, Stata, and SciPy when they apply their tie correction to arbitrary tie patterns.

## Issue: only one asymptotic convention is modelled

Grimes' simulation runs:

```r
m2 <- wilcox.test(j$group1_ranks, j$group2_ranks, correct = FALSE, exact = FALSE)
```

with `correct = FALSE`, which is the SPSS-style uncorrected asymptotic. R's default (`correct = TRUE`) applies a continuity correction; SAS, StatXact, and the SPSS Exact Tests add-on additionally use the deviation-from-mean exact convention $P(|W - \mu_W| \geq |w_\text{obs} - \mu_W|)$; Stata's `ranksum, exact` reports a mid-*p*. A reported $p$ produced by any of these conventions would not match Grimes' single asymptotic column and would therefore be incorrectly flagged inconsistent.

`{analyticgrimu}` models all four of these conventions (`p_corr_K{K}`, `p_uncorr_K{K}`, `p_exact_dev`, `p_mid`) alongside R's exact (`p_exact`).

## Issue: two-sided only

`grim_u.r` calls `wilcox.test()` without specifying an `alternative`, so every comparison is two-sided. One-sided reports cannot be checked against Grimes' candidate set, and the question of which group is labelled "1" (R vs SPSS vs SciPy convention) is not addressed.

`{analyticgrimu}::grimu_check()` accepts `alternative ∈ {"two.sided", "less", "greater"}` and, for the one-sided cases, enumerates both directions and OR's the resulting consistency checks, so that an opposite-labelling report still passes.

## Issue: non-deterministic Monte Carlo backend

Grimes' `simrank()` (and the `simrank_optimised()` variant in the Jan 2025 update of `grim_u.r`) explores random combinations of ranks until one matches the target rank sum. Without successive approximation, an "inconsistent" verdict at the end of `max_iter` iterations has to be interpreted as "the reported $p$ is impossible OR we didn't sample long enough to find it". This undermines the forensic interpretation of the test.

`{analyticgrimu}` is fully analytic: there is no Monte Carlo step, so an "inconsistent" verdict means *no candidate exists*, period (subject to the tie-pattern and asymptotic-convention coverage above).
