---
title: "Checking Mann-Whitney U results with GRIM-U"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Checking Mann-Whitney U results with GRIM-U}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(analyticgrimu)
```

## The idea

GRIM ("Granularity-Related Inconsistency of Means") checks whether a reported
mean could have come from an integer-valued measure and a stated sample size.
GRIM-U, proposed by Heathers and Grimes
([2025](https://medicalevidenceproject.org/grim-u-observation-establish-impossible-p-values-ranked-tests/)),
extends the same logic to the Mann-Whitney U test.

For two groups of sizes `n1` and `n2`, the U statistic can take only a
discrete set of values (integers, or half-integers when ties are present), and
each attainable U maps to a discrete set of p-values. A reported statistic or
p-value is *consistent* only if some attainable value falls inside its rounding
window. When none does, the reported result is mathematically impossible for
that design: it is *inconsistent*.

`analyticgrimu` performs this check while accounting for (a) the rounding of the
reported inputs and (b) the several p-value conventions used by different
statistical software.

## Checking a single result

`grimu_check()` is the main entry point. Supply the two group sizes, the
reported p-value, and the number of decimals it was reported to. The U statistic
is optional.

```{r}
res <- grimu_check(n1 = 8, n2 = 8, p_reported = 0.017, digits = 3)

res$summary[, c("n1", "n2", "p_reported", "consistent",
                "p_granularity_consistent")]
```

The `summary` is a one-row tibble with the overall verdict (`consistent`) and
the component checks. The `details` element lists every candidate U value in the
search window and which p-value convention (if any) matched.

If a U statistic is also reported, the check triangulates all three quantities:
the U must be attainable, the p-value must be attainable, and the specific U must
be able to produce the specific p-value.

```{r}
res_u <- grimu_check(n1 = 10, n2 = 12, u_reported = 30,
                     p_reported = 0.041, digits = 3)

res_u$summary[, c("consistent", "u_bounds_consistent",
                  "u_granularity_consistent", "u_matches_p")]
```

Here the reported U is a possible value and the p-value is a possible value, but
they cannot arise *together*, so the overall verdict is inconsistent.

## Checking a U statistic on its own

`check_u_consistency()` runs just the sample-size / U-statistic part: is the U
within `[0, n1 * n2]`, and is it an integer (no ties) or half-integer (ties)?

```{r}
check_u_consistency(n1 = 10, n2 = 12, u_reported = 30)
check_u_consistency(n1 = 10, n2 = 12, u_reported = 30.3)
```

## The attainable p-value space

`grimu_map_pvalues()` enumerates every attainable U value and the p-values it
implies under each convention the package models: the exact doubled-tail
(R / SciPy) value, the deviation-from-mean value (SPSS Exact / StatXact), mid-p,
and the tie-corrected asymptotic approximations.

```{r}
map <- grimu_map_pvalues(n1 = 8, n2 = 9)
head(map[, c("U", "is_integer", "p_exact")])
```

`grimu_saturation()` summarises the granularity of a design as the proportion of
all possible rounded p-values that the test can actually attain. A low value
means most p-values in the range are impossible, so a GRIM-U check has a low
false-positive rate for that design.

```{r}
grimu_saturation(n1 = 8, n2 = 9, decimals = 3)
```

## Visualising granularity

`plot_grimu_probability()` draws the attainable p-value space for one or more
sample-size pairs. Black cells mark attainable (exact-test, no-ties) p-values;
grey cells mark impossible ones.

```{r, fig.width = 5, fig.height = 6}
plot_grimu_probability(n1_vec = c(8, 30), n2_vec = c(9, 31))
```

## A caution on exact tests with ties

Some tools (`coin::wilcox_test(distribution = "exact")`, SAS
`PROC NPAR1WAY EXACT`, StatXact, the SPSS Exact Tests add-on) compute the exact
permutation p-value *conditional on the observed tie pattern*, which is not
recoverable from the sample sizes alone. An inconsistent verdict against a report
from one of those tools may reflect this coverage gap rather than a genuine
error, so check the source software before treating a flag as evidence of a
mistake.
